---
title: "Tidy Tuesday: Fifa Audience"
author: "Ellis Hughes"
output: rmarkdown::github_document
editor_options: 
  chunk_output_type: console
---

```{r dl_dataset, include=FALSE}
library(tidyverse)
fifa_audience<-read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/week11_fifa_audience.csv")
```

## First Look
The data provided for this [#TidyTuesday](https://github.com/rfordatascience/tidytuesday) is provided via [FiveThirtyEight.com](https://fivethirtyeight.com/features/how-to-break-fifa/).

```{r initial_look}
str(fifa_audience)
```

It looks like this dataset is actually pretty small, with stats for 191 countries and 6 variables including country name. 

## Initial Thoughts

The data shows what percent of the total viewership that watched the 2010 FIFA world cup. My first thought is what would be a key indicator for viewership. That would likely be rankings of your countries teams for the past 5 years. I will extract that information directly from the FIFA website, as well as the historical average ranking. Not all the country names matched between the two datasets. I attempted to match best I could figure. I apologize for any mistakes.

```{r Rankings, echo=TRUE, warning=FALSE}
library(rvest)

extractRankings<-function(url,country){
  
  urlhtml<-read_html(url)
  
  Rankings <- urlhtml%>%
    html_nodes(".col-xs-12")%>%
    html_nodes(".tbl-ranking")%>%
    html_table()%>%as.data.frame()%>%
    select(Rank,Date)
  
  Recent_Rankings<-Rankings%>%
    filter(Date<=2010, Date >= 2005)%>%
    {
      tmp<-rep(NA,6)
      names(tmp)<-paste0("Ranking_",2005:2010)
      if(nrow(.)>0){
        tmp[which(names(tmp)%in%paste0("Ranking_",.$Date))]<-.$Rank
      }
      return(tmp)
      }%>%
    as.data.frame%>%t
  
  Historical_Ranking<-Rankings%>%
    filter(Date<=2010)%>%
    data.frame%>%
    `$`("Rank")%>%
    as.numeric%>%
    mean(na.rm = TRUE)
  
  data.frame(Country=country,Recent_Rankings,Historical_Ranking=Historical_Ranking)
}

available_countries<-read_html("https://www.fifa.com/fifa-world-ranking/associations/")%>%
  html_nodes(".ranking-teamlist") %>%
  html_nodes("a") %>%
  {data.frame(Country=html_text(.),link=html_attr(.,"href"), stringsAsFactors = FALSE)}%>%
  mutate(url=paste0("https://www.fifa.com",link))%>%
  select(Country,url)%>%
  mutate(Country=dplyr::recode(Country,
                               USA="United States",
                               England="United Kingdom",
                               "China PR"="China",
                               "Korea Republic"="South Korea"
                               ))

available_countries%>%
  filter(Country %in% fifa_audience$country)%>%dim

rankings<-available_countries%>%
  filter(Country %in% fifa_audience$country)%>%
  split(., seq(nrow(.)))%>%
  map_df(function(.x){
    rankings<-try(extractRankings(.x$url,.x$Country))
    if(inherits(rankings,"try-error")){
      data.frame(Country=.x$Country,NA,NA,NA,NA,NA,NA,NA)
    }else{
      rankings
    }})%>%
  rename(country=Country)

str(rankings)

```

Not every country was included in the FIFA rankings for every country, so we are left with `r nrow(rankings)`. Doing a quick linear regression between the years will give us an indicator which ranking is most significant for predicting viewership.

```{r Modeling, echo=FALSE, warning=FALSE}
library(car)

FIFA<-merge(fifa_audience,rankings,by="country")

rankingsLM<-lm(tv_audience_share~Historical_Ranking+Ranking_2010+Ranking_2009+Ranking_2008+Ranking_2007+Ranking_2006+Ranking_2005, data=FIFA)

Anova(rankingsLM, type="III")

```

Surprisingly, it looks like the previous year, and historical ranking are the most significant for predicting viewership for the world cup! 2009 rankings were the next most significant (but not very significant), and I found that 2009 rankings were used for the seeding of the world cup after some investigation online (wikipedia) and clicking on a link to [FIFA.com](https://web.archive.org/web/20111225043322/http://www.fifa.com/worldcup/archive/southafrica2010/finaldraw/news/newsid%3D1142262/). Considering that the world cup takes place partway though the 2010 year (June-July), the 2009 rankings could provide some insight.

## Visualization

I thought the most interesting way to show viewership would be comparing historical vs 2009 ranking, and setting the size of the points based on the population, and color by viewership share

```{r visualization_1}

library(ggrepel)


FIFA_PLOT<-ggplot(FIFA)+
  geom_point(aes(x=Historical_Ranking,y=Ranking_2009,color=tv_audience_share,size=population_share))+
  geom_abline(slope=1, intercept=0, color="red")+
  scale_x_reverse()+scale_y_reverse()+
  geom_label_repel(aes(label=country,x=Historical_Ranking,y=Ranking_2009),
                   data=FIFA[which(FIFA$tv_audience_share>3 | FIFA$population_share>5),],
                   arrow = arrow(length = unit(0.03, "npc"), type = "closed", ends = "last"),
                   force = 100,
                   nudge_x = 40)+
  ggtitle("FIFA TV Viewer Share by Country")


FIFA_PLOT
```

Interestingly, most of the countries with the large populations and viewership underperformed in the 2009 year versus their historical success. 

```{r save_vis, echo=FALSE}
png("FIFA_Viewship.PNG",width = 1000,height=1000)
FIFA_PLOT
dev.off()
  
```

